---
title: "Beer Preferences for Thursday Night Football: A Blanced Incomplete Block Design Experiment"
author:
  - name: Matthew Murnane
  - name: Ethan Newcomb
format:
  pdf:
    mainfont: "Times New Roman"
indent: true
fontsize: "12"
---

```{r include=FALSE}
library(tidyverse)
library(ggthemes)
library(tidyr)
library(knitr)

beers <- c("Sierra(Pale_ale)", "Coors(Generic)", "Guinness(Dark)", "Pliny(IPA)")
Nolan <- c(NA, 6, 5.5, 3)
Jon <- c(5.5, 6, 3.5, NA)
Beni <- c(7, NA, 4, 8.5)
Zach <- c(2.5, 6.5, NA, 2)

raw_data <- data.frame(beers, Nolan, Jon, Beni, Zach)

pivoted_raw_data <- pivot_longer(raw_data,
                        cols=-beers,
                        names_to = "names",
                        values_to = "rating")

cleaned_data <- pivoted_raw_data %>% 
  drop_na(rating)
```

\newpage

\begin{center}
\textbf{\Large Abstract}
\end{center}



\newpage

\begin{center}
\textbf{\Large Introduction}
\end{center}



\newpage

\begin{center}
\textbf{\Large Methods}
\end{center}



\newpage

\begin{center}
\textbf{\Large Results}
\end{center}

```{r echo=FALSE}
raw_data %>%
  kable(
    caption = "Beer Ratings by Participant",
    col.names = c("Beers", "Nolan", "Jon", "Beni", "Zach"),
    align = "c"
  )
```

```{r echo=FALSE}
cleaned_data %>% 
  ggplot(aes(x = rating,
             y = beers))+
  geom_point(size = 2)+
  theme_few()+
  ggtitle("Ratings Per Beer")+
  ylab("Type of Beer")+
  xlab("Rating")
```

```{r echo=FALSE}
cleaned_data %>% 
  ggplot(aes(x = rating,
             y = names))+
  geom_point(size = 2)+
  theme_few()+
  ggtitle("Rating Per Subject")+
  xlab("Subjects")+
  ylab("Rating")
```




\newpage

\begin{center}
\textbf{\Large Conclusion}
\end{center}



\newpage

\begin{center}
\textbf{\Large Appendix}
\end{center}

# Code Used

### Libraries Used

```{r eval=FALSE}
library(tidyverse)
library(ggthemes)
library(tidyr)
library(knitr)
```

### Data Code

```{r eval=FALSE}
# Data input
beers <- c("Sierra(Pale_ale)", "Coors(Generic)", "Guinness(Dark)", "Pliny(IPA)")
Nolan <- c(NA, 6, 5.5, 3)
Jon <- c(5.5, 6, 3.5, NA)
Benni <- c(7, NA, 4, 8.5)
Zach <- c(2.5, 6.5, NA, 2)

raw_data <- data.frame(beers, Nolan, Jon, Benni, Zach)

# Data Cleaning

pivoted_raw_data <- pivot_longer(raw_data,
                        cols=-beers,
                        names_to = "names",
                        values_to = "rating")

cleaned_data <- pivoted_raw_data %>% 
  drop_na(rating)
```

```{r eval=FALSE}
#Table
raw_data %>%
  kable(
    caption = "Beer Ratings by Participant",
    col.names = c("Beers", "Nolan", "Jon", "Beni", "Zach"),
    align = "c"
  )
```

### Plots
Code for Plot 1 {.anchor #plot1}
```{r eval=FALSE}
cleaned_data %>% 
  ggplot(aes(x = rating,
             y = beers))+
  geom_point(size = 2)+
  theme_few()+
  ggtitle("Ratings Per Beer")+
  ylab("Type of Beer")+
  xlab("Rating")
```

```{r eval=FALSE}
cleaned_data %>% 
  ggplot(aes(x = rating,
             y = names))+
  geom_point(size = 2)+
  theme_few()+
  ggtitle("Rating Per Subject")+
  xlab("Subjects")+
  ylab("Rating")
```

